First Order Alternation

### 2018-11-17 quantifiers/body.tex
ok

Many results in formal language theory rely on the assumption that
languages are defined over finite alphabets. In practice, this
assumption is not realistic when attempting to model real-time
hardware and communication protocols or even simple programs, whose
input and observable output requires taking into account data values,
ranging over very large domains, better viewed as infinite
mathematical abstractions.

@@

The main contribution of this paper is an effective emptiness checking
semi-algorithm for first-order alternating automata, in the spirit of
the \impact~procedure, originally developed for checking safety of
nondeterministic integer programs \cite{McMillan06}. Essentially,
checking emptiness of first-order alternating automata by lazy
annotation with interpolants faces two problems: \begin{compactenum}
\item Quantified transition rules make it hard, or even impossible, to
  decide if a given symbolic trace is spurious. This is mainly because
  adding uninterpreted predicate symbols to decidable first-order
  theories, such as Presburger arithmetic, results in undecidability
  \cite{Halpern91}. To deal with this problem, we assume that the
  first order data theory, without uninterpreted predicate symbols,
  has a quantifier elimination procedure, that instantiates quantifers
  with effectively computable \emph{witness terms}.
%
\item The interpolants that prove the spuriousness of a symbolic path
  are not \emph{local}, as they may refer to input values encountered
  in the past. However, the future executions are oblivious to
  \emph{when} these values have been seen in the past and depend only
  on the data constraints between the values. We use this fact to
  define a labeling of nodes, visited by the lazy annotation
  procedure, with conjunctions of existentially quantified
  interpolants combining predicate atoms with data constraints.
\end{compactenum}

@@

This solves an older language inclusion problem
$\bigcap_{i=1}^n\lang{A_i} \subseteq \lang{B}$, between finite-state
automata with data variables, whose languages are alternating
sequences of input events and variable valuations
\cite{IosifRV16}. There, we assumed that all variables of the observer
automaton $B$ must be declared in the automata $A_1, \ldots, A_n$ that
model the concurrent components of the system under check.

@@

The execution semantics of automata is given in close analogy with the
case of boolean alternating automata, with transition rules of the
form $q \arrow{a}{} \phi$, where $q$ is a boolean constant and $\phi$
a positive boolean combination of such constants. For instance, $q_0
\arrow{a}{} q_1 \wedge q_2 \vee q_3$ means that the automaton can
choose to transition in either both $q_1$ and $q_2$ or in $q_3$
alone. This intuition leads to saying that the steps of the automaton
are defined by the minimal boolean models of the transition
formulae. In this case, both $\set{q_1 \leftarrow \top, q_2 \leftarrow
  \top, q_3 \leftarrow \bot}$ and $\set{q_1 \leftarrow \bot, q_2
  \leftarrow \bot, q_3 \leftarrow \top}$ are minimal models, however
$\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow \top}$,
is a model but is not minimal. The original definition of alternating
finite-state automata \cite{ChandraKozenStockmeyer81} works around
this problem by considering boolean valuations (models) instead of
formulae.

Given a predicate $q \in Q$ and a tuple of data values
$d_1,\ldots,d_{\#(q)}$, the tuple $q(d_1,\ldots,d_{\#(q)})$ is called
a \emph{configuration}\footnote{Note that a configuration is not a
  logical term since data values cannot be written in logic.}. To
formalize the execution semantics of automata, we relate sets of
configurations to models of first order sentences, as follows. Given a
formula $\phi$ and a valuation of its free variables $\nu$, each
interpretation $\I \in \sem{\phi}_\nu$ corresponds to a set of
configurations $\cube{\I} \isdef \set{q(d_1,\ldots,d_{\#(q)}) \mid q
  \in Q,~ \tuple{d_1,\ldots,d_{\#(q)}} \in q^\I}$, called a
\emph{cube}. For a set $\mathcal{S}$ of interpretations, we define
$\cube{\mathcal{S}} \isdef \set{\cube{\I} \mid \I \in
  \mathcal{S}}$.

@@

Given a sequence of input events $\alpha = a_1 \ldots a_n$, the
\emph{path formula} of $\alpha$ is defined as:
\begin{equation}\label{eq:pathform}
\pathform{\alpha} \isdef \stamp{\iota}{0} \wedge 
\bigwedge_{i=1}^n \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{}
  \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
\stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i}
\end{equation}
The automaton $\A$, to which $\pathform{\alpha}$ refers, will always
be clear from the context. To formalize the relation between the
low-level configuration-based execution semantics and the symbolic
path formulae, consider a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$. Any
execution forest $\T$ of $\A$ over $w$ is associated an interpretation
$\I_{\T}$ of the set of time-stamped predicates $\stamp{Q}{\leq n}$,
defined as:
\[\I_{\T}(\stamp{q}{i}) \isdef \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid
  q(d_1,\ldots,d_{\#(q)}) \text{ labels a node on level $i$ in $\T$}} 
,~\forall q\in Q ~\forall i \in [n]\]

@@

As an immediate consequence, one can decide whether $\A$ accepts some
word $w$ with a given input sequence $\event{w}=\alpha$, by checking
whether $\accform{\alpha}$ is satisfiable. However, unlike
non-alternating infinite-state models of computation, such as counter
automata (nondeterministic programs with integer variables), the
satisfiability query for an acceptance (path) formula falls outside of
known decidable theories, supported by standard SMT solvers. There are
basically two reasons for this, namely \begin{inparaenum}[(i)]
\item the presence of predicate symbols, and
\item the non-trivial alternation of quantifiers.
\end{inparaenum}
To understand this point, consider for example, the decidable theory
of Presburger arithmetic \cite{Presburger29}. Adding even only one
monadic predicate symbol to it yields undecidability in the presence
of non-trivial quantifier alternation \cite{Halpern91}. However the
quantifier-free fragment of Presburger arithmetic extended with
predicate symbols can be shown to be decidable, using a Nelson-Oppen
style congruence closure argument \cite{NelsonOppen80}.

@@

Finally, we define the acceptance of a word with a given input event
sequence by means of a formula in which no predicate atom occurs. As
previously discussed, several decidable theories, such as Presburger
arithmetic, become undecidable if predicate atoms are added to
them. Therefore, the result below makes a step towards deciding
whether the automaton accepts a word with a given input sequence, by
reducing this problem to the satisfiability of a quantified formula
without predicates.

@@

An unfolding node $\alpha \in \dom(U)$ is said to be \emph{spurious}
if and only if $\accform{\alpha}$ is unsatisfiable. In this case, we
change (refine) the labels of (some of the) prefixes of $\alpha$ (and
that of $\alpha$), such that $U(\alpha)$ becomes $\bot$, thus
indicating that there is no real execution of the automaton along that
input event sequence. As a result of the change of labels, if a node
$\gamma \prefix \alpha$ used to cover another node from $\dom(U)$, it
might not cover it with the new label. Therefore, the coverage
relation has to be recomputed after each refinement of the
labeling. The semi-algorithm stops when (and if) a safe complete
unfolding has been found. For a detailed presentation of the emptiness
procedure, we refer the reader to \cite{IosifXu18}.

@@

Typically, when checking the unreachability of a set of program
configurations \cite{McMillan06}, the interpolants used to annotate
the unfolded control structure are assertions about the values of the
program variables in a given control state, at a certain step of a
given execution. However, in an alternating model of computation, it
is useful to distinguish between
\begin{inparaenum}[(i)]
\item locality of interpolants w.r.t. a given control state (control
  locality) and
%
\item locality w.r.t. a given time stamp (time locality).
\end{inparaenum} 
In logical terms, \emph{control-local} interpolants are defined by
formulae involving a single predicate symbol, whereas
\emph{time-local} interpolants involve only predicates $\stamp{q}{i}$
and variables $\stamp{x}{i}$, for a single $i \geq 0$.

@@

\begin{theorem}\label{thm:skolem}
  Given $Q_1 x_1 \ldots Q_n x_n ~.~ \phi$ a first order sentence,
  where $Q_1, \ldots, Q_n \in \set{\exists,\forall}$ and $\phi$ is
  quantifier-free, let $\eta_i \isdef f_i(y_1,\ldots,y_{k_i})$ if $Q_i
  = \forall$ and $\eta_i \isdef x_i$ if $Q_i = \exists$, where $f_i$
  is a fresh function symbol and $\set{y_1, \ldots, y_{k_i}} =
  \set{x_j \mid j < i,~ Q_j = \exists}$. Then the entailment \(Q_1 x_1
  \ldots Q_n x_n ~.~ \phi \models \phi[\eta_1/x_1,\ldots,\eta_n/x_n]\)
  holds.
\end{theorem}
\proof{See \cite[Theorem 2.1.8]{BorgerGraedelGurevich97} and
  \cite[Lemma 2.1.9]{BorgerGraedelGurevich97}. \qed}

@@

A function $\xi : \nat \rightarrow \nat$ is [strictly]
\emph{monotonic} iff for each $n < m$ we have $\xi(n) \leq \xi(m)$
     [$\xi(n) < \xi(m)$] and \emph{finite-range} iff for each $n \in
     \nat$ the set $\set{m \mid \xi(m)=n}$ is finite. If $\xi$ is
     finite-range, we denote by $\xi_{\max}^{-1}(n) \in \nat$ the
     maximal value $m$ such that $\xi(m)=n$. It is easy to see that
     $\xi_{\max}^{-1}$ is strictly monotonic if $\xi$ is monotonic and
     finite-range. The lemma below gives the proof of correctness for
     the previous construction of non-local GLI.

@@

  \quantformn{\alpha_n}$ where it occurs. Then the following
  hold: \begin{compactenum}
  \item\label{it1:subst-quant} one can effectively build witness terms
    $\tau_{i_1}, \ldots, \tau_{i_\ell}$, where \(\set{i_1,
    \ldots,i_\ell} = \set{j \in [1,m] \mid Q_j = \forall}\) and
    \(\voc{\tau_{i_j}} \subseteq \stamp{X}{\leq \xi(i_j)} \cup
    \set{x_k \mid k < i_j, Q_k = \exists},~ \forall j \in [1,\ell]\)
    such that $\widehat{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,

@@

  \item\label{it2:subst-quant} one can effectively build a GLI $(I_0,
    \ldots, I_n)$ for $\alpha$, such that \(\voc{I_k} \subseteq
    \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup \set{x_j \mid j <
      \xi^{-1}(k),~ Q_j = \exists},~ \forall k \in [n]\).

@@

In conclusion, under two assumptions about the first order theory of
the data domain, namely the \begin{inparaenum}[(i)]
\item witness-producing quantifier elimination, and
\item Lyndon interpolation for the quantifier-free fragment with
  uninterpreted functions,
\end{inparaenum}
we developped a rather generic method that produces generalized Lyndon
interpolants for unfeasible input event sequences. Moreover, each
formula $I_k$ in the interpolant refers only to the current predicate
symbols $\stamp{Q}{I_k}$, the current and past input variables
$\stamp{X}{\leq k}$ and the existentially quantified transition
variables introduced at the previous steps $\set{x_j \mid j <
  \xi_{\max}^{-1}(k), Q_j = \exists}$. The remaining question is how
to use such non-local interpolants to label the unfolding of an
automaton (Definition \ref{def:unfolding}) and to compute the coverage
between nodes of the unfolding (Definition \ref{def:coverage}).

>>>>

Many results in formal language theory rely on the assumption that
languages are defined over finite alphabets. In practice, this
assumption is problematic when attempting to use automata as models of
real-time systems or even simple programs, whose input and observable
output requires taking into account data values, ranging over very
large domains, better viewed as infinite mathematical abstractions.

@@

The main contribution of this paper is an effective emptiness checking
semi-algorithm for first-order alternating automata, in the spirit of
the \impact~procedure, originally developed for checking safety of
nondeterministic integer programs \cite{McMillan06}. However, checking
emptiness of first-order alternating automata by lazy annotation with
interpolants faces two problems: \begin{compactenum}
\item Quantified transition rules make it hard, or even impossible, to
  decide if a given symbolic trace is spurious. This is mainly because
  adding uninterpreted predicate symbols to decidable first-order
  theories, such as Presburger arithmetic, results in undecidability
  \cite{Halpern91}. To deal with this problem, we assume that the
  first order data theory, without uninterpreted predicate symbols,
  has a quantifier elimination procedure, that instantiates quantifers
  with effectively computable \emph{witness terms}.
%
\item The interpolants that prove the spuriousness of a symbolic path
  are not \emph{local}, as they may refer to input values encountered
  in the past. However, the future executions are oblivious to
  \emph{when} these values have been seen in the past and depend only
  on the data constraints between the values. We use this fact to
  define a labeling of nodes, visited by the lazy annotation
  procedure, with conjunctions of existentially quantified
  interpolants combining predicate atoms with data constraints.
\end{compactenum}

@@

This solves an older language inclusion problem
$\bigcap_{i=1}^n\lang{A_i} \subseteq \lang{B}$, between finite-state
automata with data variables, whose languages are alternating
sequences of input events and variable valuations
\cite{IosifRV16}. There, we assumed that all variables of the observer
automaton $B$ must be declared in the automata $A_1, \ldots, A_n$ that
model the concurrent components of the system under check. Using
first-order alternating automata allows to bypass this limitation of
our previous work.

@@

The execution semantics of automata is given in close analogy with the
case of boolean alternating automata, with transition rules of the
form $q \arrow{a}{} \phi$, where $q$ is a boolean constant and $\phi$
a positive boolean combination of such constants. For instance, $q_0
\arrow{a}{} q_1 \wedge q_2 \vee q_3$ means that the automaton can
choose to transition in either both $q_1$ and $q_2$ or in $q_3$
alone. This intuition leads to saying that the steps of the automaton
are defined by the minimal boolean models of the transition
formulae. In this case, both $\set{q_1 \leftarrow \top, q_2 \leftarrow
  \top, q_3 \leftarrow \bot}$ and $\set{q_1 \leftarrow \bot, q_2
  \leftarrow \bot, q_3 \leftarrow \top}$ are minimal models, however
$\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow \top}$,
is a model but is not minimal. The original definition of alternating
finite-state automata \cite{ChandraKozenStockmeyer81} works around
this problem by considering boolean valuations (models) instead of
formulae. However, describing first-order alternating automata using
interpretations instead of formulae would be rather hard to follow.

Given a predicate $q \in Q$ and a tuple of data values
$d_1,\ldots,d_{\#(q)}$, the tuple $q(d_1,\ldots,d_{\#(q)})$ is called
a \emph{configuration}\footnote{Note that a configuration is not a
  logical term since data values cannot be written in logic.}. To
formalize the execution semantics of automata, we relate sets of
configurations to models of first order sentences, as follows. Each
first-order interpretation $\I$ corresponds to a set of configurations
$\cube{\I} \isdef \set{q(d_1,\ldots,d_{\#(q)}) \mid q \in Q,~
  \tuple{d_1,\ldots,d_{\#(q)}} \in q^\I}$, called a \emph{cube}. For a
set $\mathcal{S}$ of interpretations, we define $\cube{\mathcal{S}}
\isdef \set{\cube{\I} \mid \I \in \mathcal{S}}$.

@@

Given a sequence of input events $\alpha = a_1 \ldots a_n \in
\Sigma^*$, the \emph{path formula} of $\alpha$ is:
\begin{equation}\label{eq:pathform}
\pathform{\alpha} \isdef \stamp{\iota}{0} \wedge 
\bigwedge_{i=1}^n \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{}
  \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
\stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i}
\end{equation}
The automaton $\A$, to which $\pathform{\alpha}$ refers, will always
be clear from the context. To formalize the relation between the
low-level configuration-based execution semantics and the symbolic
path formulae, consider a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n) \in
\Sigma[X]^*$. Any execution forest $\T$ of $\A$ over $w$ is associated
an interpretation $\I_{\T}$ of the set of time-stamped predicates
$\stamp{Q}{\leq n}$, defined as:
\[\I_{\T}(\stamp{q}{i}) \isdef \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid
  q(d_1,\ldots,d_{\#(q)}) \text{ labels a node on level $i$ in $\T$}} 
,~\forall q\in Q ~\forall i \in [n]\]

@@

As an immediate consequence, one can decide whether $\A$ accepts some
word $w$ with a given input sequence $\event{w}=\alpha$, by checking
whether $\accform{\alpha}$ is satisfiable. However, unlike
non-alternating infinite-state models of computation, such as counter
automata (nondeterministic programs with integer variables), the
satisfiability query for an acceptance (path) formula falls outside of
known decidable theories, supported by standard SMT solvers. There are
basically two reasons for this, namely\begin{inparaenum}[(i)]
\item the presence of predicate symbols, and
\item the non-trivial alternation of quantifiers.
\end{inparaenum}
To understand this point, consider for example, the decidable theory
of Presburger arithmetic \cite{Presburger29}. Adding even only one
monadic predicate symbol to it yields undecidability in the presence
of non-trivial quantifier alternation \cite{Halpern91}. However the
quantifier-free fragment of Presburger arithmetic extended with
predicate symbols can be shown to be decidable, using a Nelson-Oppen
style congruence closure argument \cite{NelsonOppen80}.

@@

Finally, we define the acceptance of a word with a given input event
sequence by means of a formula in which no predicate atom occurs. As
previously discussed, several decidable theories, such as Presburger
arithmetic, become undecidable if predicate atoms are added to
them. Therefore, the result below makes a step forward towards
deciding whether the automaton accepts a word with a given input
sequence, by reducing this problem to the satisfiability of a
quantified formula without predicates.

@@

An unfolding node $\alpha \in \dom(U)$ is said to be \emph{spurious}
if and only if $\accform{\alpha}$ is unsatisfiable. In this case, we
change (refine) the labels of (some of the) prefixes of $\alpha$ (and
that of $\alpha$), such that $U(\alpha)$ becomes $\bot$, thus
indicating that there is no real execution of the automaton along that
input event sequence. As a result of the change of labels, if a node
$\gamma \prefix \alpha$ used to cover another node from $\dom(U)$, it
might not cover it with the new label. Therefore, the coverage
relation has to be recomputed after each refinement of the
labeling. The semi-algorithm stops when (and if) a safe complete
unfolding has been found. For a detailed presentation of the emptiness
procedure, we refer to \cite{IosifXu18}.

@@

Typically, when checking the unreachability of a set of program
configurations \cite{McMillan06}, the interpolants used to annotate
the unfolded control structure are assertions about the values of the
program variables in a given control state, at a certain step of an
execution. However, in an alternating model of computation, it is
useful to distinguish between
\begin{inparaenum}[(i)]
\item locality of interpolants w.r.t. a given control state (control
  locality) and
%
\item locality w.r.t. a given time stamp (time locality).
\end{inparaenum} 
In logical terms, \emph{control-local} interpolants are defined by
formulae involving a single predicate symbol, whereas
\emph{time-local} interpolants involve only predicates $\stamp{q}{i}$
and variables $\stamp{x}{i}$, for a single $i \geq 0$.

@@

\begin{theorem}[\cite{BorgerGraedelGurevich97}]\label{thm:skolem}
  Given $Q_1 x_1 \ldots Q_n x_n ~.~ \phi$ a first order sentence,
  where $Q_1, \ldots, Q_n \in \set{\exists,\forall}$ and $\phi$ is
  quantifier-free, let $\eta_i \isdef f_i(y_1,\ldots,y_{k_i})$ if $Q_i
  = \forall$ and $\eta_i \isdef x_i$ if $Q_i = \exists$, where $f_i$
  is a fresh function symbol and $\set{y_1, \ldots, y_{k_i}} =
  \set{x_j \mid j < i,~ Q_j = \exists}$. Then the entailment \(Q_1 x_1
  \ldots Q_n x_n ~.~ \phi \models \phi[\eta_1/x_1,\ldots,\eta_n/x_n]\)
  holds.
\end{theorem}
\proof{See \cite[Theorem 2.1.8]{BorgerGraedelGurevich97} and
  \cite[Lemma 2.1.9]{BorgerGraedelGurevich97}. \qed}

@@

A function $\xi : \nat \rightarrow \nat$ is [strictly]
\emph{monotonic} iff for each $n < m$ we have $\xi(n) \leq \xi(m)$
     [$\xi(n) < \xi(m)$] and \emph{finite-range} iff for each $n \in
     \nat$ the set $\set{m \mid \xi(m)=n}$ is finite. If $\xi$ is
     finite-range, we denote by $\xi_{\max}^{-1}(n) \in \nat$ the
     maximal value $m$ such that $\xi(m)=n$. The lemma below gives the
     proof of correctness for the construction of non-local GLI.

@@

  \quantformn{\alpha_n}$ where it occurs. Then one can effectively build:
  \begin{compactenum}
  \item\label{it1:subst-quant} witness terms $\tau_{i_1}, \ldots,
    \tau_{i_\ell}$, where \(\set{i_1, \ldots,i_\ell} = \set{j \in
      [1,m] \mid Q_j = \forall}\) and \(\voc{\tau_{i_j}} \subseteq
    \stamp{X}{\leq \xi(i_j)} \cup \set{x_k \mid k < i_j, Q_k =
      \exists},~ \forall j \in [1,\ell]\) such that

@@

  \item\label{it2:subst-quant} a GLI $(I_0, \ldots, I_n)$ for
    $\alpha$, such that \(\voc{I_k} \subseteq \stamp{Q}{k} \cup
    \stamp{X}{\leq k} \cup \set{x_j \mid j < \xi^{-1}(k),~ Q_j =
      \exists}\), for all $k \in [n]$. 

@@

In conclusion, under two assumptions about the first order theory of
the data domain, namely the\begin{inparaenum}[(i)]
\item witness-producing quantifier elimination, and
\item Lyndon interpolation for the quantifier-free fragment with
  uninterpreted functions,
\end{inparaenum}
we developped a rather generic method that produces generalized Lyndon
interpolants for unfeasible input event sequences. Moreover, each
formula $I_k$ in the interpolant refers only to the current predicate
symbols $\stamp{Q}{I_k}$, the current and past input variables
$\stamp{X}{\leq k}$ and the existentially quantified transition
variables introduced at the previous steps $\set{x_j \mid j <
  \xi_{\max}^{-1}(k), Q_j = \exists}$. The remaining question is how
to use such non-local interpolants to label the unfolding of an
automaton (Definition \ref{def:unfolding}) and to compute the coverage
between nodes of the unfolding (Definition \ref{def:coverage}).
